---
title: "Persistent Kinase and Chromatin Memory in LPS-Tolerized Human Monocytes"
subtitle: "Integrated RNA-seq and ChIP-seq re-analysis of GSE85246"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    df_print: paged
    self_contained: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "95%"
)

options(stringsAsFactors = FALSE)

safe_read <- function(path) {
  if (!file.exists(path)) {
    stop(paste("Required file not found:", path))
  }
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

safe_include <- function(path, width = "95%") {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(paste0("**Figure not found:** `", path, "`"))
  }
}

fmt_num <- function(x, digits = 2) {
  ifelse(is.na(x), NA, formatC(x, digits = digits, format = "f"))
}

# Core result tables
day6 <- safe_read(
  "results/03_memory_kinases/03_memory_kinases_FDR05.csv"
)

integrated <- safe_read(
  "results/05_integrated_trajectory/05_integrated_kinase_trajectory.csv"
)

chromatin <- safe_read(
  "results/07b_unbiased_distal_chromatin/07b_corrected_gene_level_chromatin_summary.csv"
)

top <- safe_read(
  "results/08_final_chromatin/08_TOP_integrated_candidates.csv"
)

# Reproducibility status is optional for the narrative, but useful if present.
status_file <- "results/run_all_status.csv"
status <- if (file.exists(status_file)) {
  read.csv(status_file, stringsAsFactors = FALSE, check.names = FALSE)
} else {
  NULL
}

# Known final counts are derived directly from output tables wherever possible.
n_memory <- nrow(integrated)

n_high <- if ("RNA_chromatin_priority" %in% names(integrated)) {
  sum(integrated$RNA_chromatin_priority == "HIGH", na.rm = TRUE)
} else {
  NA_integer_
}

n_chr_any <- if ("Any_corrected_chromatin_support" %in% names(chromatin)) {
  sum(chromatin$Any_corrected_chromatin_support, na.rm = TRUE)
} else {
  NA_integer_
}

n_chr_dual <- if ("Any_corrected_dual_mark_support" %in% names(chromatin)) {
  sum(chromatin$Any_corrected_dual_mark_support, na.rm = TRUE)
} else {
  NA_integer_
}

high_chr <- chromatin[
  chromatin$RNA_chromatin_priority == "HIGH",
  ,
  drop = FALSE
]

n_high_chr_any <- if ("Any_corrected_chromatin_support" %in% names(high_chr)) {
  sum(high_chr$Any_corrected_chromatin_support, na.rm = TRUE)
} else {
  NA_integer_
}

n_high_chr_dual <- if ("Any_corrected_dual_mark_support" %in% names(high_chr)) {
  sum(high_chr$Any_corrected_dual_mark_support, na.rm = TRUE)
} else {
  NA_integer_
}
```

# Executive Summary

This report presents a downstream computational re-analysis of the human monocyte innate-memory dataset **GSE85246**, integrating RNA-seq state, β-glucan rescue/restimulation behavior, and chromatin signal from **H3K27ac** and **H3K4me1** tracks.

The analysis identified `r n_memory` Day-6 memory-associated kinase candidates. Integrated RNA evidence prioritized `r n_high` kinases as HIGH-priority candidates. After correcting the distal chromatin analysis to avoid post-hoc best-window selection, `r n_chr_any` of 24 memory kinases showed directional chromatin support and `r n_chr_dual` showed corrected dual-mark support. Among the 10 HIGH RNA-priority kinases, `r n_high_chr_any` showed corrected chromatin support and `r n_high_chr_dual` showed corrected dual-mark support.

The final integrated candidates were:

```{r top-symbols, results='asis'}
cat(
  paste0(
    "**",
    paste(top$SYMBOL, collapse = ", "),
    "**"
  )
)
```

These candidates combine persistent Day-6 RNA-state evidence, rescue/restimulation support, and corrected chromatin evidence. The chromatin component is interpreted as **directional and descriptive** rather than proof of causal epigenetic inheritance.

# 1. Biological Question

The biological question was whether transient LPS exposure leaves a persistent kinase-associated molecular state after washout, whether β-glucan treatment can reverse or remodel that state, and whether persistent transcriptional changes are accompanied by persistent regulatory chromatin changes.

The analysis therefore followed the sequence:

**acute response → post-washout memory → kinase prioritization → rescue/restimulation → integrated RNA trajectory → chromatin persistence → final RNA/chromatin integration**

This ordering was chosen to distinguish initial LPS responsiveness from the later persistent memory state.

# 2. Dataset and Experimental Structure

The project analyzes data from **GSE85246**, including RNA-seq and ChIP-seq components associated with LPS tolerance and β-glucan-mediated reprogramming in primary human monocytes/macrophages.

The main analytical comparisons were:

- **Day 1:** LPS versus RPMI, representing the initial response.
- **Day 6:** LPS-history versus RPMI-history after washout, representing the persistent state.
- **Restimulation:** comparison of restimulation responses between LPS-history and RPMI-history cells.
- **β-glucan rescue:** directional assessment of whether rescue moved expression or chromatin signal back toward the RPMI state.
- **Chromatin:** H3K27ac and H3K4me1 at predefined promoters and condition-blind selected non-promoter loci.

# 3. Analytical Workflow

The canonical reproducible pipeline is:

```text
01_setup_and_data.R
        ↓
02_initial_LPS_response.R
        ↓
03_persistent_kinase_memory.R
        ↓
04_BG_rescue_and_restimulation.R
        ↓
05_integrated_kinase_trajectory.R
        ↓
06_make_final_kinase_figure.R
        ↓
07_epigenetic_persistence.R
        ↓
07b_unbiased_distal_chromatin.R
        ↓
08_make_final_chromatin_figures.R
```

The complete pipeline can be rerun from the project root using:

```bash
Rscript --vanilla run_all.R
```

Each analysis step is executed in a separate clean R process.

# 4. Day-1 Initial LPS Response

Day-1 LPS-versus-RPMI analysis was used to characterize the acute response rather than define the final memory-kinase candidate set.

Because only two matched Day-1 pairs were available, the acute analysis had limited multiple-testing power. Day-1 effects were therefore used later as supporting trajectory information rather than as a mandatory filter for memory-candidate selection.

```{r fig-day1-day6}
safe_include(
  "figures/06_final_kinase/06C_Day1_vs_Day6_kinase_effects.png"
)
```

This design allows a kinase to be retained as a genuine Day-6 memory candidate even when its Day-1 signal is weak, thereby avoiding the assumption that every persistent state must first show a strong acute transcriptional response.

# 5. Day-6 Persistent Kinase Memory

The primary kinase-memory set was defined directly from the Day-6 LPS-history versus RPMI-history comparison using the predefined canonical kinase family.

The final Day-6 memory set contained **24 kinases**.

```{r day6-table}
show_cols <- intersect(
  c(
    "SYMBOL",
    "log2FC",
    "P.Value",
    "adj.P.Val",
    "kinase_FDR"
  ),
  names(day6)
)

if (length(show_cols) >= 2) {
  knitr::kable(
    day6[, show_cols, drop = FALSE],
    caption = "Day-6 memory-associated kinase candidates"
  )
} else {
  knitr::kable(
    day6,
    caption = "Day-6 memory-associated kinase candidates"
  )
}
```

```{r fig-day6}
safe_include(
  "figures/06_final_kinase/06B_Day6_memory_kinase_effects.png"
)
```

The candidate-family FDR is interpreted conditionally within the predefined kinase family, while genome-wide evidence is retained separately for transparency.

# 6. β-Glucan Rescue and Restimulation

Restimulation analysis tested whether the response to a new LPS challenge differed according to previous LPS exposure.

The formal matched response model compared:

\[
(LPS_{restim} - LPS_{baseline}) -
(RPMI_{restim} - RPMI_{baseline})
\]

No memory kinase reached genome-wide significance for the history-by-restimulation interaction after multiple-testing correction. Consequently, interaction results were not treated as independent statistical confirmation.

Instead, rescue/restimulation evidence was integrated directionally:

- whether β-glucan baseline expression moved toward RPMI;
- whether the restimulation response was blunted or altered;
- whether β-glucan rescue moved the restimulation response toward the naive/RPMI-like response.

Because the rescue arm had only two samples, rescue conclusions are descriptive and directional.

# 7. Integrated RNA Trajectory

The RNA-level integration retained all 24 formal Day-6 memory kinases while assigning an evidence-based follow-up priority.

```{r fig-rna-integrated}
safe_include(
  "figures/06_final_kinase/06D_FINAL_integrated_kinase_summary.png"
)
```

The HIGH-priority RNA set contained **10 kinases**:

```{r high-rna-list, results='asis'}
high_symbols <- integrated$SYMBOL[
  integrated$RNA_chromatin_priority == "HIGH"
]

cat(paste(high_symbols, collapse = ", "))
```

Priority is an integrated directional-evidence label and **not a new significance test**.

# 8. Chromatin Persistence Analysis

Chromatin analysis used normalized BigWig signal for:

- H3K27ac
- H3K4me1

The analysis examined both:

- a predefined promoter window around each gene TSS;
- a non-promoter window selected without using the LPS/RPMI/BG direction.

For a mark to count as directional persistent/reversible support, it had to satisfy:

1. Day-1 LPS−RPMI and Day-6 LPS−RPMI had the same direction;
2. the Day-6 rescue shift opposed the Day-6 LPS−RPMI direction;
3. rescue was closer to RPMI than the Day-6 LPS state.

# 9. Correction of Distal-Window Selection Bias

The first exploratory distal implementation scanned many 1-kb windows around each candidate gene and then selected the most favorable window. This produced an implausibly high rate of dual-mark support and created a post-hoc selection problem.

The corrected analysis therefore selected **one fixed non-promoter window per gene using condition-blind chromatin abundance**, before testing any Day-1, Day-6, or rescue direction.

This corrected analysis produced:

- **15/24** promoter any-mark support;
- **6/24** promoter dual-mark support;
- **14/24** fixed non-promoter any-mark support;
- **3/24** fixed non-promoter dual-mark support;
- **18/24** any corrected chromatin support;
- **7/24** any corrected dual-mark support.

Among the 10 HIGH RNA candidates:

- **8/10** showed corrected chromatin support;
- **3/10** showed corrected dual-mark support.

```{r fig-chromatin-all}
safe_include(
  "figures/08_final_chromatin/08A_corrected_chromatin_support_all24.png"
)
```

# 10. Final RNA + Chromatin Integration

```{r fig-final-integration}
safe_include(
  "figures/08_final_chromatin/08D_FINAL_RNA_chromatin_integration.png"
)
```

The final three leading integrated candidates are:

```{r top-table}
top_cols <- intersect(
  c(
    "SYMBOL",
    "Day6_memory_log2FC",
    "Day6_memory_FDR_kinase",
    "Directional_support_score",
    "Promoter_mark_support_count",
    "Distal_mark_support_count",
    "Corrected_RNA_plus_chromatin_class"
  ),
  names(top)
)

knitr::kable(
  top[, top_cols, drop = FALSE],
  caption = "Final integrated candidates"
)
```

# 11. MAP3K8

MAP3K8 emerged as a HIGH-priority RNA candidate with positive Day-6 memory expression and integrated rescue/restimulation support.

Its corrected chromatin pattern was strongest at the promoter:

- promoter H3K27ac: supported;
- promoter H3K4me1: supported;
- fixed non-promoter H3K27ac: supported;
- fixed non-promoter H3K4me1: not supported.

IGV inspection showed real chromatin structure at the locus and a pattern broadly compatible with the computational result. The most defensible interpretation is that MAP3K8 has **strong promoter-centered dual-mark chromatin support**, with additional H3K27ac support at the fixed non-promoter locus.

# 12. BMPR1A

BMPR1A showed a negative Day-6 RNA memory effect and strong integrated directional support.

It had the broadest corrected chromatin support among the final trio:

- promoter H3K27ac: supported;
- promoter H3K4me1: supported;
- fixed non-promoter H3K27ac: supported;
- fixed non-promoter H3K4me1: supported by the predefined directional criteria.

The H3K27ac fixed-locus pattern was particularly strong, with rescue returning the average signal close to the RPMI state. H3K4me1 also showed persistent/reversal direction, but the rescue response overshot RPMI and was therefore interpreted more cautiously.

Overall, BMPR1A has particularly strong integrated chromatin evidence, especially for H3K27ac.

# 13. JAK3

JAK3 showed a positive Day-6 RNA memory effect and HIGH integrated RNA priority.

The promoter showed dual-mark support:

- promoter H3K27ac: supported;
- promoter H3K4me1: supported.

At the fixed non-promoter/proximal region:

- H3K27ac: supported;
- H3K4me1: not supported because rescue did not become closer to RPMI.

Because the selected non-promoter window lies immediately outside the operational promoter boundary, it is best described as a **proximal non-promoter regulatory window**, not as a proven distal enhancer.

# 14. Directional Chromatin Patterns in the Final Candidates

```{r fig-top-direction}
safe_include(
  "figures/08_final_chromatin/08C_TOP_candidates_chromatin_direction.png"
)
```

The figure is internally scaled within each mark-region unit. Therefore, it should be interpreted for **direction and pattern**, not for absolute comparison of effect size between H3K27ac and H3K4me1.

# 15. IGV Validation

IGV inspection was used as a final qualitative validation step for the three leading candidates.

The hg19 loci used were generated automatically by Step 08 and stored in:

```text
results/08_final_chromatin/08_TOP_candidate_IGV_loci_hg19.csv
```

IGV was used to verify that the selected promoter/non-promoter regions corresponded to genuine local chromatin signal and that the visual direction was broadly compatible with the quantified analysis.

IGV validation is supportive rather than statistical.

# 16. Limitations

The main limitations are:

1. **Day-1 RNA sample size:** only two matched pairs were available, limiting FDR power for the acute-response analysis.
2. **β-glucan rescue sample size:** rescue RNA analysis had only two samples and is therefore directional/descriptive.
3. **Restimulation interaction:** no memory kinase reached genome-wide FDR significance for the history-by-restimulation interaction.
4. **ChIP donor structure:** Day-1 and Day-6 ChIP samples are not longitudinally paired donors.
5. **Chromatin inference:** same-locus Day-1/Day-6 signal does not prove causal inheritance of a histone modification.
6. **Operational regulatory windows:** the non-promoter regions are computationally selected windows, not experimentally validated enhancers.
7. **Downstream re-analysis:** the RNA workflow is a downstream re-analysis of MMSEQ estimates using limma and is not an exact reproduction of the original publication's MMDIFF differential-expression pipeline.

# 17. Conclusion

This re-analysis identifies a persistent kinase-associated transcriptional state following LPS exposure and washout. Twenty-four canonical kinases met the Day-6 memory criteria, and integration with β-glucan rescue/restimulation evidence prioritized ten at the RNA level.

After correcting the chromatin workflow to avoid post-hoc distal-window selection, three candidates—**MAP3K8, BMPR1A, and JAK3**—showed the strongest combined RNA and dual-mark chromatin evidence.

The results support these genes as high-value candidates for mechanistic follow-up. However, the chromatin patterns should be interpreted as **persistent, same-locus directional associations** rather than proof that H3K27ac or H3K4me1 causally transmits cellular memory.

# 18. Reproducibility

```{r run-status}
if (!is.null(status)) {
  knitr::kable(
    status[
      ,
      intersect(
        c(
          "step",
          "label",
          "status",
          "elapsed_minutes",
          "exit_status"
        ),
        names(status)
      ),
      drop = FALSE
    ],
    caption = "Canonical run-all pipeline status"
  )
} else {
  cat("`results/run_all_status.csv` was not found.")
}
```

The canonical analysis can be reproduced with:

```bash
Rscript --vanilla run_all.R
```

The final pipeline is considered complete when all nine canonical steps report `SUCCESS`.

# 19. Key Output Files

```text
results/05_integrated_trajectory/05_integrated_kinase_trajectory.csv
figures/06_final_kinase/06D_FINAL_integrated_kinase_summary.png

results/07b_unbiased_distal_chromatin/
  07b_corrected_gene_level_chromatin_summary.csv

results/08_final_chromatin/
  08_TOP_integrated_candidates.csv
  08_TOP_candidate_IGV_loci_hg19.csv
  08_FINAL_professor_table.csv

figures/08_final_chromatin/
  08A_corrected_chromatin_support_all24.png
  08B_HIGH_RNA_Day6_effect_with_chromatin_class.png
  08C_TOP_candidates_chromatin_direction.png
  08D_FINAL_RNA_chromatin_integration.png
```

# 20. Data Source

Primary dataset: **GSE85246**

Associated components used in this project include RNA-seq and ChIP-seq data from the LPS tolerance / β-glucan reprogramming study by Novakovic and colleagues.

