---
title: "TCGA Pan-Cancer Kinase Exploration"
subtitle: "Integrated expression, clinical and immune-infiltration analyses"
---

```{r}
availability <- read.csv(
  "tables/JAK3_complete_33cancer_analysis_availability.csv",
  check.names = FALSE
)

flags <- read.csv(
  "tables/JAK3_master_key_flags_corrected.csv",
  check.names = FALSE
)

surv <- read.csv(
  "tables/JAK3_TCGA_CDR_survival_complete.csv",
  check.names = FALSE
)
```

## Scope

This report provides a standardized TCGA pan-cancer framework for
candidate kinase evaluation.

For each TCGA cancer type, analyses are included where the required
data are available:

- tumor versus normal expression
- matched tumor-normal expression
- early-stage tumor versus normal expression
- stage-wise expression trends
- grade-wise expression trends
- prevalence of high or aberrant expression
- survival associations as secondary evidence
- sample-size and data-availability limitations
- immune/inflammatory infiltration controls

The completed analysis now contains two linked modules:

- **JAK3** — the reference pan-cancer kinase analysis
- **Multikinase** — comparative RNA, clinical and ATAC-RNA analysis of the expanded candidate panel

The multikinase module uses JAK3 as the completed benchmark and extends the
framework to nine additional candidates emerging from the innate immune-memory
analysis.

## JAK3 analysis coverage

```{r}
coverage <- data.frame(
  Analysis = c(
    "TCGA projects evaluated",
    "Tumor versus normal",
    "Matched tumor-normal",
    "Stage-I tumor versus normal",
    "Stage trend",
    "Grade trend",
    "Prevalence analysis",
    "Survival analysis"
  ),
  Cancer_projects = c(
    nrow(availability),
    sum(availability$Tumor_vs_normal_available, na.rm = TRUE),
    sum(availability$Matched_available, na.rm = TRUE),
    sum(availability$Early_stage_vs_normal_available, na.rm = TRUE),
    sum(availability$Stage_trend_available, na.rm = TRUE),
    sum(availability$Grade_trend_available, na.rm = TRUE),
    sum(availability$Prevalence_available, na.rm = TRUE),
    sum(surv$Analysis_status == "Analysed", na.rm = TRUE)
  )
)

knitr::kable(
  coverage,
  align = c("l", "r")
)
```

## Current JAK3 summary

```{r}
summary_tab <- data.frame(
  Result = c(
    "Canonical tumor-normal significant UP",
    "Canonical Stage-I-normal significant UP",
    "Unadjusted OLS tumor-normal positive",
    "Positive association persisting after PTPRC adjustment",
    "Positive association persisting after 10-marker ImmuneScore adjustment",
    "Unadjusted OLS Stage-I positive",
    "Stage-I positive association persisting after PTPRC adjustment",
    "Stage-I positive association persisting after 10-marker ImmuneScore adjustment"
  ),
  Cancer_projects = c(
    sum(flags$Canonical_TN_significant_UP, na.rm = TRUE),
    sum(flags$Canonical_StageI_significant_UP, na.rm = TRUE),
    sum(flags$TN_OLS_unadjusted_positive, na.rm = TRUE),
    sum(flags$TN_positive_persists_after_PTPRC, na.rm = TRUE),
    sum(flags$TN_positive_persists_after_ImmuneScore, na.rm = TRUE),
    sum(flags$StageI_OLS_unadjusted_positive, na.rm = TRUE),
    sum(flags$StageI_positive_persists_after_PTPRC, na.rm = TRUE),
    sum(flags$StageI_positive_persists_after_ImmuneScore, na.rm = TRUE)
  )
)

knitr::kable(
  summary_tab,
  align = c("l", "r")
)
```

## Data availability across 33 TCGA projects

```{r}
availability_display <- availability[
  ,
  c(
    "project",
    "Primary_malignant_patients",
    "Normal_patients",
    "Matched_pairs",
    "Stage_I_patients",
    "Patients_with_stage",
    "Patients_with_grade",
    "Tumor_vs_normal_available",
    "Matched_available",
    "Early_stage_vs_normal_available",
    "Stage_trend_available",
    "Grade_trend_available",
    "Survival_available"
  )
]

knitr::kable(
  availability_display,
  align = "l"
)
```

## Interpretation

TCGA RNA-seq represents bulk tissue. Observed kinase expression may
therefore reflect both malignant-cell biology and differences in
immune or stromal cell content.

Associations that remain after adjustment for measured immune proxies
are described as **immune-adjusted or residual associations**. They
should not be interpreted as proof of malignant-cell-intrinsic
expression.

Early-stage elevation is considered evidence of an association with
early disease and does not establish a causal role in tumor initiation.

## Report sections

The complete JAK3 analysis is provided on the **JAK3** page, including:

- tumor-normal comparisons
- matched-pair analysis
- Stage-I analysis
- stage and grade trends
- high/aberrant-expression prevalence
- survival
- leukocyte-fraction analysis
- CIBERSORT analysis
- PTPRC/CD45 adjustment
- multi-marker immune-score adjustment
- integrated 33-cancer results

Detailed analytical definitions are provided on the **Methods** page.

The **Multikinase** page provides the comparative extension, including
patient-matched ATAC-RNA regulatory analysis and final multi-omic candidate
comparison.
